---
title: "Data Analysis Workflow"
output:
---
# Step 0 Set working directory
```{r}
getwd()
setwd()
```
**Relative** - `setwd("./data")`, `setwd("../")` (Move up one directory)
**Absolute** - `setwd("/Users/jtleek/data/")` "/" for path in windows
## Checking for and creating directories
```{r}
# check to see if the directory exists
file.exists("directoryName") 
# create a directory if it doesn't exist
dir.create("directoryName") 

# an example checking for a "data" directory and creating it if it doesn't exist
if(!file.exists("data")){
  dir.create("data")
}
```
# Step 1 Get data
## From the internet
```{r}
# Download the file
fileUrl <- "https://data.baltimorecity.gov/api/views/dz54-2aru/rows.csv?accessType=DOWNLOAD"
download.file(fileUrl,destfile="./data/cameras.csv",mode = 'wb')

# Check the downloaded file
list.files("./data")
# Record the date downloaded
dateDownloaded <- date()
dateDownloaded
```
## From local flat files
### read.table
```{r}
cameraData <- read.table("./data/cameras.csv",sep=",",header=TRUE)
```
Important parameters
- *file*
- *header*
- *sep* - delimiter
- *row.names*
- *nrows* - how many rows to read of the file (e.g. nrows=10 reads 10 lines)
- *quote* - whether there are any quoted values, quote="" means no quotes
- *na.strings* - set the character that represents a missing value
- *skip* - number of lines to skip before starting to read
### read.csv
```{r}
cameraData <- read.csv("./data/cameras.csv")
```
`read.csv` sets *sep=","* and *header=TRUE*
