---
title: "R Programming"
---
# 1. Pipes
The pipe, `%>%`, comes from the `magrittr` package by Stefan Milton Bache. Packages in the tidyverse load `%>%` for you automatically
```{r}
library(magrittr)
```
## Other tools from magrittr
- `%T>%` (“tee” pipe) works like `%>%` except that it returns the left-hand side instead of the right-hand side.
```{r}
rnorm(100) %>%
  matrix(ncol = 2) %>%
  plot() %>%
  str()
#>  NULL

rnorm(100) %>%
  matrix(ncol = 2) %T>%
  plot() %>%
  str()
#>  num [1:50, 1:2] -0.387 -0.785 -1.057 -0.796 -1.756 ...
```
- If you’re working with functions that don’t have a data frame based API
(i.e. you pass them individual vectors, not a data frame and expressions to be evaluated in the context of that data frame), you might find `%$%` useful. It “explodes” out the variables in a data frame so that you can refer to them explicitly. This is useful when working with many functions in base R:
```{r}
mtcars %$%
  cor(disp, mpg)
#> [1] -0.848
```
# 2. Functions
## Conditional execution
use `dplyr::near()` for comparisons
`switch()` allows you to evaluate selected code based on position or name.
```{r}
cut(temp, c(-Inf, 0, 10, 20, 30, Inf), right = TRUE,
    labels = c("freezing", "cold", "cool", "warm", "hot"))
```
## Function arguments
useful compromise is the built-in `stopifnot()`: it checks that each argument is TRUE, and produces a generic error message if not.
```{r}
wt_mean <- function(x, w, na.rm = FALSE) {
  stopifnot(is.logical(na.rm), length(na.rm) == 1)
  stopifnot(length(x) == length(w))
  
  if (na.rm) {
    miss <- is.na(x) | is.na(w)
    x <- x[!miss]
    w <- w[!miss]
  }
  sum(w * x) / sum(w)
}
```
## Return values
There are two basic types of pipeable functions: transformations and side-effects. With transformations, an object is passed to the function’s first argument and a modified object is returned. With side-effects, the passed object is not transformed. Instead, the function performs an action on the object, like drawing a plot or saving a file. Side-effects functions should “invisibly” return the first argument, so that while they’re not printed they can still be used in a pipeline. For example, this simple function prints the number of missing values in a data frame:
```{r}
show_missings <- function(df) {
  n <- sum(is.na(df))
  cat("Missing values: ", n, "\n", sep = "")
  
  invisible(df)
}

x <- show_missings(mtcars) 
class(x)
dim(x)

mtcars %>% 
  show_missings() %>% 
  mutate(mpg = ifelse(mpg < 20, NA, mpg)) %>% 
  show_missings()
```
# 3. Iteration
## For loop variations
### Unknown output length
A better solution to save the results in a list, and then combine into a single vector after the loop is done:
```{r}
means <- c(0, 1, 2)
out <- vector("list", length(means))
for (i in seq_along(means)) {
  n <- sample(100, 1)
  out[[i]] <- rnorm(n, means[[i]])
}
str(out)
str(unlist(out))
```
Here I’ve used `unlist()` to flatten a list of vectors into a single vector. A stricter option is to use `purrr::flatten_dbl()` — it will throw an error if the input isn’t a list of doubles.
This pattern occurs in other places too:
1. You might be generating a long string. Instead of `paste()`ing together each iteration with the previous, save the output in a character vector and then combine that vector into a single string with `paste(output, collapse = "")`.
2. You might be generating a big data frame. Instead of sequentially `rbind()`ing in each iteration, save the output in a list, then use `dplyr::bind_rows(output)` to combine the output into a single data frame.
Watch out for this pattern. Whenever you see it, switch to a more complex result object, and then combine in one step at the end.
## The map functions
- `map()` makes a list.
- `map_lgl()` makes a logical vector.
- `map_int()` makes an integer vector.
- `map_dbl()` makes a double vector.
- `map_chr()` makes a character vector.
Each function takes a vector as input, applies a function to each piece, and then returns a new vector that’s the same length (and has the same names) as the input. The type of the vector is determined by the suffix to the map function.
```{r}
map_dbl(df, mean)
map_dbl(df, median)
map_dbl(df, sd)

df %>% map_dbl(mean)
df %>% map_dbl(median)
df %>% map_dbl(sd)
```
- `map_*()` uses … ([dot dot dot]) to pass along additional arguments to `.f` each time it’s called:
```{r}
map_dbl(df, mean, trim = 0.5)
```
- The map functions also preserve names:
```{r}
z <- list(x = 1:3, y = 4:5)
map_int(z, length)
#> x y 
#> 3 2
```
### Shortcuts
The second argument, `.f`, the function to apply, can be a **formula**, a **character vector**, or an **integer vector**.
The syntax for creating an anonymous function in R is quite verbose so purrr provides a convenient shortcut: a **one-sided formula**.
```{r}
models <- mtcars %>% 
  split(.$cyl) %>% 
  map(~lm(mpg ~ wt, data = .))
```
Here I’ve used `.` as a pronoun: it refers to the current list element (in the same way that i referred to the current index in the for loop).
When you’re looking at many models, you might want to extract a summary statistic like the R-square. To do that we need to first run `summary()` and then extract the component called `r.squared`. We could do that using the shorthand for anonymous functions:
```{r}
models %>% 
  map(summary) %>% 
  map_dbl(~.$r.squared)
#>     4     6     8 
#> 0.509 0.465 0.423
```
But extracting named components is a common operation, so purrr provides an even shorter shortcut: you can use a **string**.
```{r}
models %>% 
  map(summary) %>% 
  map_dbl("r.squared")
```
You can also use an **integer** to select elements by position:
```{r}
x <- list(list(1, 2, 3), list(4, 5, 6), list(7, 8, 9))
x %>% map_dbl(2)
#> [1] 2 5 8
```
## Deal with failure
`safely()` is an adverb: it takes a function (a verb) and returns a modified version. It always returns a list with two elements:
1. `result` is the original result. If there was an error, this will be `NULL`.
2. `error` is an error object. If the operation was successful, this will be `NULL`.
```{r}
safe_log <- safely(log)
str(safe_log(10))
#> List of 2
#>  $ result: num 2.3
#>  $ error : NULL
str(safe_log("a"))
#> List of 2
#>  $ result: NULL
#>  $ error :List of 2
#>   ..$ message: chr "non-numeric argument to mathematical function"
#>   ..$ call   : language log(x = x, base = base)
#>   ..- attr(*, "class")= chr [1:3] "simpleError" "error" "condition"
```

```{r}
x <- list(1, 10, "a")
y <- x %>% map(safely(log))
str(y)
```
This would be easier to work with if we had two lists: one of all the errors and one of all the output. That’s easy to get with `purrr::transpose()`:
```{r}
y <- y %>% transpose()
str(y)
```
typically you’ll either look at the values of `x` where `y` is an error, or work with the values of `y` that are ok:
```{r}
is_ok <- y$error %>% map_lgl(is_null)
x[!is_ok]
y$result[is_ok] %>% flatten_dbl()
```
`purrr` provides two other useful adverbs:
- Like `safely()`, `possibly()` always succeeds. It’s simpler than `safely()`, because you give it a default value to return when there is an error.
```{r}
x <- list(1, 10, "a")
x %>% map_dbl(possibly(log, NA_real_))
#> [1] 0.0 2.3  NA
```
- `quietly()` performs a similar role to safely(), but instead of capturing errors, it captures printed output, messages, and warnings:
```{r}
x <- list(1, -1)
x %>% map(quietly(log)) %>% str()
```
## Map over multiple arguments
`map2()` iterates over two vectors in parallel:
```{r}
map2(mu, sigma, rnorm, n = 5) %>% str()
#> List of 3
#>  $ : num [1:5] 4.78 5.59 4.93 4.3 4.47
#>  $ : num [1:5] 10.85 10.57 6.02 8.82 15.93
#>  $ : num [1:5] -1.12 7.39 -7.5 -10.09 -2.7
```
Note that the arguments that vary for each call come before the function; arguments that are the same for every call come after.  
`pmap()` takes a list of arguments. You might use that if you wanted to vary the mean, standard deviation, and number of samples:
```{r}
n <- list(1, 3, 5)
args1 <- list(n, mu, sigma)
args1 %>%
  pmap(rnorm) %>% 
  str()
#> List of 3
#>  $ : num 4.55
#>  $ : num [1:3] 13.4 18.8 13.2
#>  $ : num [1:5] 0.685 10.801 -11.671 21.363 -2.562
```
If you don’t name the elements of list, `pmap()` will use positional matching when calling the function. That’s a little fragile, and makes the code harder to read, so it’s better to name the arguments:
```{r}
args2 <- list(mean = mu, sd = sigma, n = n)
args2 %>% 
  pmap(rnorm) %>% 
  str()
```
Since the arguments are all the same length, it makes sense to store them in a data frame:
```{r}
params <- tribble(
  ~mean, ~sd, ~n,
    5,     1,  1,
   10,     5,  3,
   -3,    10,  5
)
params %>% 
  pmap(rnorm)
```
As soon as your code gets complicated, a data frame is a good approach because it ensures that each column has a name and is the same length as all the other columns.
### Invoke different functions
`invoke_map()`: vary the function
```{r}
f <- c("runif", "rnorm", "rpois")
param <- list(
  list(min = -1, max = 1), 
  list(sd = 5), 
  list(lambda = 10)
)
invoke_map(f, param, n = 5) %>% str()
```
The first argument is a list of functions or character vector of function names. The second argument is a list of lists giving the arguments that vary for each function. The subsequent arguments are passed on to every function.  
And again, you can use `tribble()` to make creating these matching pairs a little easier:
```{r}
sim <- tribble(
  ~f,      ~params,
  "runif", list(min = -1, max = 1),
  "rnorm", list(sd = 5),
  "rpois", list(lambda = 10)
)
sim %>% 
  mutate(sim = invoke_map(f, params, n = 10)) %>% 
  str()
```
## Walk
Walk is an alternative to map that you use when you want to call a function for its side effects, rather than for its return value. You typically do this because you want to render output to the screen or save files to disk - the important thing is the action, not the return value.
```{r}
x <- list(1, "a", 3)

x %>% 
  walk(print)
#> [1] 1
#> [1] "a"
#> [1] 3
```
`walk()` is generally not that useful compared to `walk2()` or `pwalk()`. For example, if you had a list of plots and a vector of file names, you could use `pwalk()` to save each file to the corresponding location on disk:
```{r}
library(ggplot2)
plots <- mtcars %>% 
  split(.$cyl) %>% 
  map(~ggplot(., aes(mpg, wt)) + geom_point())
paths <- stringr::str_c(names(plots), ".pdf")

pwalk(list(paths, plots), ggsave, path = tempdir())
```
`walk()`, `walk2()` and `pwalk()` all invisibly return `.x`, the first argument. This makes them suitable for use in the middle of pipelines.
## Other patterns of for loops
### Predicate functions
A number of functions work with **predicate** functions that return either a single `TRUE` or `FALSE`.  
`keep()` and `discard()` keep elements of the input where the predicate is `TRUE` or `FALSE` respectively:
```{r}
iris %>% 
  keep(is.factor) %>% 
  str()

iris %>% 
  discard(is.factor) %>% 
  str()
```
`some()` and `every()` determine if the predicate is true for any or for all of the elements.
```{r}
x <- list(1:5, letters, list(10))

x %>% 
  some(is_character)
#> [1] TRUE

x %>% 
  every(is_vector)
#> [1] TRUE
```
`detect()` finds the first element where the predicate is true; `detect_index()` returns its position.
```{r}
x <- sample(10)
x %>% 
  detect(~ . > 5)
x %>% 
  detect_index(~ . > 5)
```
`head_while()` and `tail_while()` take elements from the start or end of a vector while a predicate is true:
```{r}
x %>% 
  head_while(~ . > 5)
x %>% 
  tail_while(~ . > 5)
```
### Reduce and accumulate
Sometimes you have a complex list that you want to reduce to a simple list by repeatedly applying a function that reduces a pair to a singleton. This is useful if you want to apply a two-table dplyr verb to multiple tables. For example, you might have a list of data frames, and you want to reduce to a single data frame by joining the elements together:
```{r}
dfs <- list(
  age = tibble(name = "John", age = 30),
  sex = tibble(name = c("John", "Mary"), sex = c("M", "F")),
  trt = tibble(name = "Mary", treatment = "A")
)

dfs %>% reduce(full_join)
```
Or maybe you have a list of vectors, and want to find the intersection:
```{r}
vs <- list(
  c(1, 3, 5, 6, 10),
  c(1, 2, 3, 7, 8, 10),
  c(1, 2, 3, 4, 8, 9, 10)
)

vs %>% reduce(intersect)
```
The reduce function takes a “binary” function (i.e. a function with two primary inputs), and applies it repeatedly to a list until there is only a single element left.  
Accumulate is similar but it keeps all the interim results. You could use it to implement a cumulative sum:
```{r}
x <- sample(10)
x %>% accumulate(`+`)
```